---
title: "**Functional Programming in `R`**"
author: "Wyclife Agumba Oluoch"
date: '`r Sys.time()`'
output: 
    html_document:
      toc: true
      toc_depth: 2
      toc_float: true
      theme: cerulean
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functional programing in `R`

Virtually every time one uses `R`, a function  is involved. It is not enough to use a function, it is enough to understand what the function does and even better still to tweak the function to suit your needs or even develop your own function(s).

Here is a function that takes a number and squares it:

```{r}
square <- function(a) a**2
square(3)
```

It can be written using curly braces as follows:

```{r}
square <- function(b){ # Create the function name
  b**2
}
square(3)
```

The same can also be written as follows:

```{r}
square <- function(c){ 
  product <- c**2
  return(product)
}
square(3)
```

The same can be presented like:

```{r}
square <- function(d){
  product <- d**2
  print(product)
}
square(5)
```

This is performing the similar task:

```{r}
square <- function(e){
  product <- e**2
  (product)
}
square(9)
```

The same can be returned using the code that follows:

```{r}
square <- function(f){
  product <- f**2
  show(product)
}
square(9)
```

Now I want the formula to return half the square root of the number:

```{r}
square_root <- function(b){
  product <- b^0.5
  return(product/2)
}
square_root(9)
```

```{r}
square_root
```

The function can also take a vector and return their outputs accordingly.

```{r}
square(-10:10)
```
or even

```{r}
square_root(c(-26, 1:15, 158))
```
We can dig into a list object and find the result we want.

```{r}
m <- list(12, 18, 345)
square_root(m[[2]])
```

# Rescale function

Here we wil to create a function that will take a vector, calculate mean, calculate standard deviation, then divide deviations of values from mean by standard deviation

```{r}
rescale <- function(x){
  m <- mean(x)
  s <- sd(x)
  (x - m) / s
}
rescale(c(-52, 58, 46, 99))
```

```{r}
rescale <- function(x, only_translate){
  m <- mean(x)
  translated <- x - m
  if (only_translate) return(translated)
  s <- sd(x)
  translated / s
}

rescale(x = 1:6, only_translate = TRUE)
```

```{r}
rescale <- function(x, only_translate){
  m <- mean(x)
  translated <- x - m
  if (only_translate) return(translated)
  s <- sd(x)
  translated / s
}

rescale(x = 1:4, only_translate = FALSE)
```

Since we have already named the arguments in the function, the order normally does not matter. That is we could start with the only_translate argument before including the x argument as below:

```{r}
rescale(only_translate = TRUE, x = 1:4)
```
In case of only two arguments being fed into a function, provided one is named, the other may not necessarily be named too. For example:

```{r}
rescale(x = 1:4, TRUE)
rescale(TRUE, x = 1:4)
rescale(1:4, only_translate = T)
rescale(only_translate = T, 1:4)
```
It is possible to give few unique initials of the name of the argument and the function will stirun successfully. For example, in our `rescale` function, the arguments start with x and o, so only stating the first name character is enough as follows:

```{r}
rescale(1:4, o = T)
rescale(o = T, 1:4)
```

The method above is suited for short codes in R and should not be used when writing programs or packages as the users may be confused. Avoid it.

# Default parameters

It is possible to give a function default parameter when defining it. This can be changed by the user when giving appropriate parameter to the argument to show results otherwise.

```{r}
rescale <- function(x, only_translate = FALSE){
  m <- mean(x)
  translated <- x - m
  if(only_translate) return(translated)
  s <- sd(x)
  translated / s
}
```


Now, since the default value of only_translate has been given, we can only supply the vector x and nothing for only_translate and the FALSE parameter for that argument will be run under the hood.

```{r}
rescale(1:4) # This runs with only_translate being taken as FALSE by default.
```

# Gobble up everything else argument

There is the `...` variable in `R` that can be used in a function. It is normally applicable when calling a function within a function and you wish to use all functions from the other function.

```{r}
rescale <- function(x, ...){
  m <- mean(x, ...)
  s <- sd(x, ...)
  (x - m) / s
}

rescale(1:4) # 
```

```{r}
x <- c(NA, 1:3)
rescale(x)
```

```{r}
x <- c(NA, 1:3)
rescale(x, na.rm = TRUE)
```
The parameter `...` takes any named parameter. However, if it is not included in a function, then you are only limited to arguments in the function and cannot add anything else like `na.rm = TRUE`.

```{r}
some_function <- function(x) x
other_function <- function(x, ...) x
# some_function(1:4, foo = "bar") # This does not run because there was no allowance.
other_function(1:4, foo = "bar")

```

```{r}
f <- function(...) list(...)
g <- function(x, y, ...) f(...)
g(x = 1, y = 2, z = 3, w = 4)
```

 Here is a function which is taking a raster stack as input and filters out those with pixel values meeting a stated threshold.
 
```{r}
compare_rasters <- function(x) {
  if (sum(x > 0.5, na.rm = TRUE) > 7) {
    return(1)
  } else {
    return(0)
  }
  
}
```
 
This function will create a binary raster layer which has pixel values of 1 in case at least 7 of the feeding raster layers from the stack have pixel value of not less than 0.5. This is an important function es when considering the aggregation of raster layers from the output of predict function in `sdm` package.
One can decide and say that if `r 3 + 4` out of the `r sqrt(169)` models successfully run are showing suitability of `r 1 - 0.3` then the pixel is considered suitable and not otherwise.

# for loop versus map loop

Here I check the difference in performance between a for loop in base R and the map loop from purrr.

```{r}
library(tidyverse)
volumes <- rnorm(1000000)

for_loop <- vector()

for (k in 1:20) {
  for_loop[k] <- system.time(for (i in 1:length(volumes)) {
    2.3 * volumes[i] ^ 9
  })
}

map_loop <- vector()
for (j in 1:20) {
  map_loop[j] <- system.time(map_vec(volumes, \(x) 2.3 * x ^ 9))
}
```

```{r}
data.frame(for_loop = for_loop, map_loop = map_loop) %>% 
  gather(key = "key", value = "value") %>% 
  ggplot(aes(key, value)) +
  geom_boxplot() +
  labs(x = "Type of loop",
       y = "Time in seconds") +
  theme(text = element_text(size = 14, face = "bold", colour = "black")) +
  theme(axis.text = element_text(size = 14, face = "bold", colour = "black"))
```
It is clear from the above plot that the for loop took less time iterating through the 1,000,000 length vector.

# References
