---
title: "<center>**Functional Programming in `R`**</center>"
author: "<center>Wyclife Agumba Oluoch</center>"
date: '<center>`r Sys.time()`</center>'
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions in `R`

Almost every time one needs to use R, a function must be involved. It is not enough to use a function, it is enough to understand what the function does and even better still tweak it to suit your needs or even develop your own function.

Here I demonstrate a function which is taking a number and squaring it:

```{r}
square <- function(a) a**2
square(3)
```

The same can be writted using the curly braces as follows:

```{r}
square <- function(b){
  b**2
}
square(4)
```

The same can also be written as:

```{r}
square <- function(b){
  product <- b**2
  return(product)
}
square(8)
```

The same can also be presented as:

```{r}
square <- function(m){
  product <- m**2
  print(product)
}
square(5)
```

This is also performing the same work:

```{r}
square <- function(k){
  product <- k**2
  (product)
}
square(9)
```

Now I want the formula to return half the square root the number:

```{r}
square_root <- function(b){
  product <- b^0.5
  return(product/2)
}
square_root(9)
```

The function can also take a list and return their outputs accordingly"

```{r}
square(1:10)
```
or even

```{r}
square_root(c(26, 1:15, 158))
```


# References
